Modularly Metaprogrammable Type Structure

### 2015-05-28 atlam-icfp15/atlam-icfp15.tex
ICFP 15

%\documentclass[12pt]{article}
\documentclass[pldi]{sigplanconf-pldi15}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\usepackage{times}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

@@

\newcommand{\moutput}{^{\color{gray}+}}
\newcommand{\rulename}[1]{(#1)}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\usepackage[compact]{titlesec}  
\titlespacing{\section}{0pt}{6pt}{2pt}
\titlespacing{\subsection}{0pt}{3pt}{0pt}
\titlespacing{\subsubsection}{0pt}{3pt}{0pt}
\titlespacing{\paragraph}{0pt}{3pt}{5pt}

@@

@@

\title{Modular Composing Typed Language Fragments}
%\subtitle{Modular Type Constructors}

\authorinfo{}{}{}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
Researchers often describe type systems as fragments, leaving to language designers the task of composing these to form complete programming  languages. 
This has not been a systematic process: metatheoretic results must be established anew for each composition of fragments, guided only informally by metatheorems derived for simpler systems.
As the language design space grows, mechanisms that provide stronger modular reasoning principles than this are needed.

In this paper, we take a foundational approach, specifying an extensible typed translation semantics, @$\lambda$. Only the $\rightarrow$ type constructor is built in; all others (we discuss constrained strings and variants of record types) are defined by extending a \emph{tycon context}. Each tycon defines associated term-level \emph{opcons} (e.g. row projection) using a static language where types and translations are values. The semantics come with modular metatheoretic guarantees, notably \emph{type safety} and \emph{conservativity}: that \emph{tycon-specific  invariants} established in any ``closed world'' are conserved in the ``open world''. Remarkably, extension providers do not need to  provide the semantics with mechanized specifications or proofs. Instead, these guarantees arise from a form of translation validation that, taking inspiration from the ML module system, uses type abstraction to check that the opcons associated with a tycon respect the \emph{translation independence} of all others.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are usually described as consisting of \emph{fragments}, each contributing separately to the concrete and abstract syntax and static and dynamic semantics of the language \cite{pfpl}. 
Harper organizes such fragments around \emph{type constructors}, e.g. $\rightarrow$ for total function types, describing each in a different chapter of his book. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language of partial (i.e. possibly non-terminating) function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Another common practice is to describe a fragment using a simple calculus having a ``catch-all'' constant and base type to stand notionally for all ``other'' terms and types.% (e.g. \cite{sanitation-psp14}).

In contrast, the usual metatheoretic reasoning techniques for programming languages  (e.g., rule induction) operate on complete language specifications. Each {combination} of fragments must formally be treated as its own monolithic language for which  metatheorems must be established anew, guided only informally by those derived for the smaller systems from which the language is notionally composed.

%In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.

This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, a list type constructor need not be built in because lists are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

%But while universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, 
Establishing an isomorphic embedding of a fragment is not always possible, however, because it requires preserving both the static and dynamic semantics and, if defined, performance bounds specified by a cost semantics, of all operators. This is possible if they have ``function-like'' semantics, as the list operators do. But in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because this requires introducing row projection operators,  written $\opname{\#}\conclbl{lbl}$ in ML, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$  is thus an \emph{operator constructor}). Each time such a situation occurs,  a new \emph{dialect} is needed. The ML lineage, like others, faces a proliferation of dialects that go beyond ``core ML'':%, showing that this is not an isolated example:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types). %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec. \ref{prior-work}).

@@

\begin{enumerate}
\vspace{-5px}
%\compresslist
\item 
\textbf{General Purpose Fragments:} 
Many variations on product types, for example, exist: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), and 
records with width and depth subtyping \cite{Cardelli:1984:SMI:1096.1098} and functional update operators\footnote{The Haskell wiki notes that ''No, extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}%, 
 %mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
%and 
% ``methods'' (i.e. pure objects) \cite{TSLs}
. 
 Sum-like types are also exposed variously: standard datatypes, 
open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
exception types \cite{Tofte:89:TheDefinitionOfStandardML}. Combinations occur as class-based object systems \cite{ocaml-manual}.

\item
\textbf{Specialized Fragments:} Fragments providing more specialized operators are also commonly introduced in dialects, e.g. for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, concurrency \cite{AliceLookingGlass}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09} and regular string sanitation \cite{sanitation-psp14}.%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.
\vspace{-3px}
\item
\textbf{Foreign Fragments:} A safe and natural foreign function interface (FFI) can be valuable. This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{enumerate}
\vspace{-5px}

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed between every pair of dialects). %And as just mentioned, composing dialects is non-trivial.

These problems with composition do not arise when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} can enforce abstraction barriers that ensure that the isomorphism needs only to be established in the ``closed world'' of the module. For example, a module for sets in ML can hold its representation type abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g. uniqueness, if using a list representation). These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.% Other languages expose \emph{abstract data types} in other ways, also to localize reasoning \cite{liskov1974programming}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

The alarming proliferation of dialects above suggests that mechanisms  that make it possible to define and  reason in a similarly modular, localized manner about  direct extensions to the semantics of a language are needed. For example, once a language is extended with \emph{regular string types} as described in \cite{sanitation-psp14}, all terms having a regular string type like $\sty{\tcvar{rstr}}{\concrx{.+}}$ should continue to behave as non-empty strings no matter which other extensions are in use. %Such a mechanism could ultimately be integrated directly into the language, blurring the distinction between fragments and libraries and decreasing the need for new dialects.% Importing fragments that introduce new semantics   would be as easy and reasonable as importing a new module is in ML today. %In the limit, the community could rely on modularly mechanized metatheory and compiler correctness results.% rather than requiring heroic efforts from individual research groups that consider an entire language at once. %Recent work has shown progress on modularly introducing new concrete syntax, reviewed in Sec. \ref{desugaring}. Our focus is on the problem of introducing new semantics.

\paragraph{Contributions} In this paper, we take foundational steps toward this goal of modular language metatheory by  constructing a simple but surprisingly powerful  calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics consist of an extensible \emph{external language} (EL) governed by a {typed translation semantics} targeting a simple and fixed \emph{internal language} (IL). 
In particular, rather than building in a monolithic set of external type and operator constructors, the  semantics are indexed by a \emph{tycon context}. Each tycon, e.g. $\tcvar{lprod}$ defining labeled products, directly specifies the semantics of its associated opcons, e.g. $\opname{\#}$ for row projection, via \emph{static functions}, i.e. functions written in a \emph{static language} (SL), where types and translations are values. %In other words, the semantics are controlled by static \emph{code and type generation}.

We call a tycon associated with opcons in this manner a \emph{modular tycon} because all \emph{translation invariants} maintained by the opcons associated with that tycon, e.g. the regular string invariant just mentioned, are necessarily maintained in any further extended tycon context, i.e. in the ``open world'', due to a remarkably simple translation validation procedure that maintains \emph{translation independence} between tycons using type abstraction in the IL. This is, encouragingly, the same fundamental principle underlying representation independence in ML-style module systems. As in ML, mechanized specifications and proofs are not needed.%, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

% We will begin by giving an overview of  @$\lambda$ in Sec. \ref{atlam}, introducing our two main examples, one defining labeled product types (i.e. record types maintaining an ordering for simplicity) with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}. In Sec. \ref{types}, we detail how tycons are defined and how types arise from tycons. We then describe how tycons control the semantics of their term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world'').  Interestingly, type system providers need not provide mechanized proofs to maintain such guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to use the same parametricity results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We conclude with  related and future work in Sec. \ref{prior-work}.

>>>>

%\documentclass[12pt]{article}
\documentclass[preprint]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

\newcommand{\Verse}{{\textsf{\small Verse}}}

\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\usepackage{times}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={module,tycon,tcsig,translation,ana,syn,intro,indexed,by,tcstruct},
literate={~} {$\sim$}{1},
numbersep=5pt,
numberstyle=\small\ttfamily\color{mygray}}

@@

\newcommand{\moutput}{^{\color{gray}-}}
\newcommand{\rulename}[1]{(#1)}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.3\baselineskip\lineskip=.3\baselineskip plus .2\baselineskip}

@@

\usepackage[compact]{titlesec}  
\titlespacing*{\section}{0pt}{0pt}{3pt}
\titlespacing*{\subsection}{0pt}{3pt}{0pt}
\titlespacing*{\subsubsection}{0pt}{3pt}{0pt}
\titlespacing*{\paragraph}{0pt}{3pt}{5pt}

@@

\newcommand{\liv}[1]{\lstinline{#1}}
\usepackage{enumitem}

@@

\title{Modularly Metaprogrammable Type Structure}
%\subtitle{Modular Type Constructors}

\authorinfo{\vspace{-10px}}{}{}
%\authorinfo{Cyrus Omar \and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %         {\{comar, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
% 
% This has not been a systematic process: metatheoretic results must be reestablished for each combination of fragments, guided only informally by metatheorems derived for simpler systems.
% As the language design space grows, mechanisms that systematize this process and provide stronger modular reasoning principles than this are needed.

We introduce \Verse, a programming language specified, like  many modern languages are, by a {typed translation semantics} targeting\- a common intermediate language (e.g. CIL bytecode, though for simplicity, we use a simply typed lambda calculus). In contrast to large such languages like Scala, the external language of {\Verse} is remarkably small, building in only function types directly. In lieu of other type structure, {\Verse} provides a metaprogramming mechanism that gives library providers the ability to modularly express type structure of their own design. For example, record types are not built in, but they can be  expressed by defining a \emph{tycon structure} that introduces a type constructor, \texttt{record}, 
%indexed by a finite mapping from static labels to types, 
together with record introduction and row projection operators. The logic governing the typechecking and translation of these operators is defined using a static metalanguage where types and translations are manipulated as values. %In other words, the mapping from the external language to the internal language is governed by metaprograms written in the static language, rather than fixed \emph{a priori}.

{\Verse} performs translation validation to guarantee type safety and \emph{conservativity}: that a broad class of metatheorems about such logic, including those establishing {translation invariants}, need only be established in a ``closed world'', i.e. as if the collection of tycon  structures was finite. These are necessarily conserved in the ``open world'', i.e. when new tycon structure are introduced, with no new proof obligations. This critical modularity result relies on type abstraction to ensure that each tycon structure maintains \emph{translation independence} with respect to all others.
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation

\section{Introduction}\label{intro}

A key feature of the ML {module system} is that it supports type abstraction at interface boundaries \cite{MacQueen:1984:MSM:800055.802036}. This enables programming ``in the large'' by localizing reasoning about data representation invariants and making it possible to  evolve these representations without breaking client programs. %By holding the representation of a type abstract outside of a module, the language guarantees that representation invariants at that type need only be established locally, in the ``closed world'' of the module. Introducing a new module into a program does not  require reestablishing such invariants. 
There is substantial agreement between dialects of ML about the value of this fundamental feature. Other languages are also increasingly converging on an ML-like design, e.g. Haskell \cite{Kilpatrick:2014:BRH} and Scala \cite{conf/oopsla/AminRO14}. \Verse, the language that we will introduce in this paper, does not buck this trend, providing a module system essentially identical to the Standard ML module system \cite{MacQueen:1984:MSM:800055.802036}. %An example is discussed in the supplement.%We make only syntactic changes, discussed in the supplement, consistent with Verse's layout-sensitive concrete syntax.

Where there are more substantial differences between languages is with regard to the semantics of the underlying core language, i.e. the language of types and expressions. Indeed, the language design community has produced hundreds of full-scale languages, toy dialects and minimal calculi that explore different points in the design space of core language constructs. For example:
%\vspace{-4px}
\begin{itemize}
\setlength{\itemsep}{-1pt}
\item 
\textbf{General Purpose Constructs:} 
Many variations on products exist: 
$n$-ary tuples, 
labeled tuples, 
records (identified up to reordering), records with extension and update operators\footnote{The Haskell wiki states that ''extensible records are not implemented in GHC. The problem is that the record design space is large, and seems to lack local optima. [...] As a result, nothing much happens.'' \cite{GHCFAQ}} \cite{ocaml-manual}, and 
records with width and depth coercions \cite{Cardelli:1984:SMI:1096.1098}%, 
 %mutable fields \cite{ocaml-manual}, 
%field delegation \cite{atlang-gpce14} \todo{gpce submission} 
%and 
% ``methods'' (i.e. pure objects) \cite{TSLs}
. 
Disjunctive types also come in variants: standard datatypes, 
open datatypes \cite{conf/ppdp/LohH06,journals/toplas/MillsteinBC04}, 
polymorphic variants \cite{ocaml-manual} and 
exception types \cite{Tofte:89:TheDefinitionOfStandardML}. Combinations occur as class-based object systems, of which there are a seemingly endless variety.

\item
\textbf{Specialized Constructs:} More specialized constructs are also commonly introduced in dialects, e.g. for distributed programming \cite{Murphy:2007:TDP:1793574.1793585}, concurrency \cite{AliceLookingGlass}, reactive programming \cite{mandel2005reactiveml}, databases \cite{Ohori:2011:MSM:2034773.2034815},  units of measure \cite{conf/cefp/Kennedy09}, typechecking XML schemas \cite{HosoyaVouillonPierce2000ICFP}, web programming \cite{conf/popl/Chlipala15} and string sanitation \cite{sanitation-psp14}.%\verb|sprintf| in the OCaml dialect statically distinguishes format strings from strings.% All of these are implemented as dialects of existing languages, presumably because a strong encoding was not feasible.

\item
l\textbf{Foreign Constructs:} A safe and natural foreign function interface (FFI) can be valuable for interacting with legacy or low-level device code. This requires enforcing the type system of the foreign language in the calling language. %Using  a FFI that does not do this can lead to safety issues, even when both languages are separately known to be safe. 
%Safe FFIs generally require direct extensions to the language. 
For example, MLj builds in a safe FFI to Java \cite{Benton:1999:IWW:317636.317791}.
\end{itemize}
\vspace{-3px}

% Such problems  do not arise for constructs that are expressed inside a language with a modern \emph{module system}, like ML, because the module system . For example, a module for sets in ML can hold its representation type abstract, ensuring that any invariants necessary to maintain the isomorphism need only be maintained by the functions in the module (e.g. uniqueness, if using a list representation). These then continue to hold no matter which other modules are in use by a client \cite{harper1997programming}.% Other languages expose \emph{abstract data types} in other ways, also to localize reasoning \cite{liskov1974programming}. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

% %In both cases, it is left as an extrinsic concern to ensure that the metatheory developed separately is actually conserved when fragments defined in these ways are composed to form a complete language.

% This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, a list type constructor need not be built in because lists are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

%But while universality properties (e.g. ``Turing completeness'') are often enough to guarantee that an embedding that preserves a desirable fragment's dynamic semantics can be constructed, 
% Establishing an isomorphic embedding of a fragment is not always possible, however, because it requires preserving both the static and dynamic semantics and, if defined, performance bounds specified by a cost semantics, of all operators. This is possible if they have ``function-like'' semantics, as the list operators do. But in $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$, it is impossible to introduce record types as a library because this requires introducing row projection operators,  written $\opname{\#}\conclbl{lbl}$ in ML, one for each of the infinite set of row labels $\conclbl{lbl}$ ($\opname{\#}$  is thus an \emph{operator constructor}). Each time such a situation occurs,  a new \emph{dialect} is needed. 
%, showing that this is not an isolated example:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%%You can only use $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ as a \emph{target of compilation} for a language with record types (they can be implemented using nested binary products,  lists, or other types). %(and attempting to simulate records via a hash table over strings plus a refinement type system to check that the lookup operation will not fail will not satisfy a cost semantics prescribing $\mathcal{O}(1)$ lookup; see Sec. \ref{prior-work}).

@@

This \emph{dialect-oriented} state of affairs is profoundly anti-modular. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
In other words, there is generally no good way to interface with a library written using a dialect different from one's own when a direct FFI is not available. There are, of course, bad ways to attempt  this. For example, one can find a compiler for the foreign dialect that targets a language for which an FFI is available, e.g. JVM or CIL bytecode, and then program against the code it generates. But this bypasses the language semantics (i.e. the interface), exposing implementation details directly. Only the type system of the intermediate language, not of the dialect itself, constrains this action. If the compiler maintains any stronger translation invariants, there is  no guarantee that they will be maintained. Moreover, if a different compiler is used, or the  compiler evolves its choices regarding data representations, client code can break. %Simply put, interacting directly with code a compiler generates is unlikely to be a healthy experience.

It is clear that the only scalable way to program ``in the large'' is to stick to a single programming language that provides strong modular reasoning principles. Once one accepts that dialects are better rhetorical vehicles than practical artifacts, there are two options. One, exemplified by languages like Scala and Ocaml, is to attempt to build in a large number of constructs into the core language \emph{a priori}, and progressively add new ones as they emerge in backwards compatible revisions. Power is centralized around a small group of language designers. This results in a large core language, but because of a ``long tail'' effect, constructs that are situationally useful, or where there is  disagreement, can still end up marginalized. Other constructs that are included but turn out to be  flawed remain entrenched (and monopolize finite ``syntactic resources'', as we will discuss below).

{\Verse} aims to explore the polar opposite end of this spectrum: a core language that builds in only minimal type structure \emph{a priori}. Instead, the core language is organized around a metaprogramming mechanism that gives library providers the ability to introduce new type structure of their own design, within reasonable constraints.

The main challenge in decentralizing the semantics of a core language  comes in maintaining  strong metatheoretic properties. For example, if we simply conceived of the core language's semantics as being specified by a ``bag of rules'' and let library providers introduce new rules, one could easily define rules that violated type safety, interfered with invariants being maintained by rules defined in other libraries, or introduced non-determinism. There is also a question of how to allocate finite syntactic resources. For example, as the footnote above noted, the design space around record types is large. If we allow libraries to explore this design space, which library gets to determine the meaning of a form like \liv{\{label1: e1, label2: e2\}}?

This paper is intended to introduce principled solutions to these problems. The key construct that the core language is organized around is called the \emph{tycon structure}. A tycon structures defines a type constructor and the semantics of the operators associated with it using statically evaluated metafunctions. The language enforces an abstraction barrier around each tycon structure that ensures that they cannot interfere with one another, which preserves modular reasoning. Encouragingly, type abstraction is critical to this abstraction barrier, much like it is to the abstraction barrier between modules. A bidirectional type system is used to resolve ambiguities with regard to common forms, like the record introduction form shown above (following recent work on ambiguity-free syntax extensions in the Wyvern programming language \cite{TSLs}).

We will discuss labeled tuples with functional update, records and \emph{regular strings} (based on a recent specification \cite{sanitation-psp14}) as examples of constructs that are, or would need to be, built in to comparable languages but that can be expressed as modular tycon structures in {\Verse}. This mechanism is not all-encompassing -- there are many compelling constructs that cannot be expressed in {\Verse} today. We will discuss its limitations more specifically in Sec. \ref{sec:discussion}. But this paper sets the ``ground rules'' of the game: we expect future versions of the language, and competing languages, to provide metatheoretic guarantees at least as strong as those we establish here, while being ever more expressive. %% This paper is intended to introduce a novel mechanism that we argue puts it uniquely on a path toward this goal.

The remainder of the paper is organized as follows:\vspace{-3px}
\begin{itemize}[noitemsep]
\item In Sec. 2, we give a tutorial-style introduction to \Verse. %We give intuition about the translation validation process that guarantees type safety and modular reasoning. 
\item In Secs. 3-4, we give a formal account of tycon structures and the translation validation process with a typed lambda calculus, $\lambda_\textsf{Verse}$. The accompanying supplement contains more details.
\item In Sec. 5, we summarize the key metatheoretic properties of the calculus introduced in Sec. 3. 
\item In Sec. 6, we compare tycon structures to other related work, and discuss present limitations and avenues for future work.
\end{itemize}

\section{\textsf{Verse} by Example}
From the perspective of an everyday programmer, {\Verse} handles very much like other typed function programming languages. For example, let us build a (toy) academic conference management application in \Verse, beginning with some type synonyms:

\vspace{-3px}\begin{lstlisting}
let type Title = rstr /.+/
let type Conf = rstr /([A-Z]+) (\d\d\d\d)/
let type Paper = record {
    title : Title,
    conf : Conf
}
\end{lstlisting}\vspace{-3px}
Starting from the bottom up, \liv{Paper} is a \emph{record type} that specifies two rows. The first is labeled \liv{title} and specifies type \liv{Title}, and the second is labeled \liv{conf} and specifies type \liv{Conf}.

The types \liv{Title} and \liv{Conf} are \emph{regular string types}, which behave as strings known statically to be in a regular language specified by a regular expression \cite{sanitation-psp14}. Here, \liv{Title} specifies non-empty strings, and \liv{Conf} specifies conference names in the standard format, i.e. a series of capital letters followed by a four-digit year. Parentheses indicate captured groups.

All types in {\Verse} arise by applying a \emph{type constructor} (\emph{tycon}, for short) to a statically valued \emph{type index}. Here, the tycon \liv{rstr} is indexed by statically valued regular expressions, and \liv{record} is indexed by statically valued finite mappings from labels to types. We are able to write the indices above using conventional concrete syntax not because this syntax is built in to {\Verse}, but because {\Verse} supports \emph{type specific languages} (TSLs), a feature recently introduced in the Wyvern programming language that permits modular reasoning about syntax extensions \cite{TSLs}. For concision, we will not go into a detailed discussion about TSLs in this paper.

We can now define a function that generates papers as follows:
\begin{lstlisting}
fun exmpl_paper(title : Title) -> Paper
  {title=title, conf="EXMPL 2015"}
\end{lstlisting}
and then call it and perform some further operations:
\begin{lstlisting}
let paper = exmpl_paper "Collapsing The Multiverse"
let year = paper#conf#2
\end{lstlisting}
The value of \liv{year} projects out the row labeled \liv{conf} from the record \liv{paper}, and then the substring corresponding to the second captured group, i.e. the year. Had we specified an incorrect row label, or an out-of-bounds captured group index, we would encounter a static type error. Note also that type annotations were not needed on these bindings, because {\Verse} supports local type inference \cite{Pierce:2000:LTI:345099.345100}. The type of \liv{year} here is synthesized to be \liv{rstr /\\d\\d\\d\\d/}.

If {\Verse} were organized like a standard language and simply built in the type structure that we saw being used above -- functions, records and regular strings -- this would be only somewhat novel (regular strings are not widely available in languages like Ocaml and Scala). But neither records nor regular strings are built directly in to {\Verse}. Instead, the core language of {\Verse} is structured like the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}. It consists of an \emph{external language} (EL) defined by a type-directed translation to a much simpler \emph{typed internal language} (IL). In other words, all external types and expressions (like those just defined) map, as they are being typechecked, to internal types and expression, respectively. Only the IL has a conventional dynamic semantics.

Only function types are built in to the EL. The {\Verse} IL, for the purposes of this paper, builds in only functions, nullary and binary products, binary sums, recursive types, strings and numbers, i.e. it is reminiscent of a simple first-stage compiler IL. In practice, we would include a few other typical constructs, e.g. reference cells, contiguous arrays and concurrency primitives, but our design is largely independent of the details of the IL, so we omit discussion of these details.

Ultimately, the translation generated for the example above will be the following (assuming suitable helper functions, not shown):
\vspace{-3px}\begin{lstlisting}
let type Title = string
let type Conf = string
let type Paper = Title * (Conf * unit)

fun exmpl_paper(title : Title) -> Paper
  (title, ("EXMPL 2015", ()))

let paper = exmpl_paper "Collapsing the Multiverse"
let year = groupPrj 2 /([A-Z]+) (\d\d\d\d)/ (
  fst(snd(paper)))
\end{lstlisting}\vspace{-3px}
External functions  translate directly to internal functions. All other types and operations, however, delegate to \emph{statically evaluated metafunctions} associated with \emph{tycon structures}. Let us consider the \liv{record} tycon structure in Figure \ref{fig:record-conc}.

The first line names the tycon structure and then ascribes a \emph{tycon signature} to it, which specifies the ``interface'' associated with the tycon structure. Here, line 2 states that the tycon is indexed by static values of \emph{kind} \liv{LblTyMap}. Kinds serve as the types of static values (discussed further in the next section). We assume that this kind has been defined previously  and specifies kind-specific syntax as mentioned when we described the record type \liv{Paper} above.
